%!TEX encoding=UTF-8 Unicode
\documentclass[a4paper,10pt]{book}
\usepackage{lmodern}
\input{custom/magic}

% first page
\usepackage[a4paper]{style/meta-donnees}

% encoding and stuff
\usepackage[utf8]{inputenc}
\usepackage[french,english]{babel}
\usepackage[T1]{fontenc}

\input{custom/macros.tex}

% Fancy style
\input{custom/fancy.tex}

% Links

\usepackage[pdfusetitle]{hyperref}

\hypersetup{
    colorlinks=true, %colore les liens
    breaklinks=true, %permet le retour à la ligne dans les liens trop longs
    urlcolor= blue, %couleur des hyperliens
    linkcolor= black, %couleur des liens internes
    bookmarksopen=true,
    citecolor=black,
}
\usepackage{url}

% Images
\usepackage[]{caption}
\usepackage[]{subcaption}
\usepackage{graphicx}
\graphicspath{{./img/}{./style/}{./tikz/}}


% Algorithms
\input{custom/algo.tex}


% Todo notes
\usepackage[obeyDraft]{todonotes}
%\usepackage{todonotes}

% Misc
\usepackage{siunitx}
\usepackage{multirow,booktabs}

\input{custom/glossary}

\input{custom/tikz}


\title{Analyzing the memory behavior of parallel scientific applications}
\author{David Beniamine}

\begin{document}

\frontmatter

\input{custom/firstpage.tex}

\startcontents[smalltoc]

\begin{chapter}{Abstract}

    Since a few decades, high performances computers or more and more parallel.
    At the same, time the gap between processors and memory frequency increased significantly.
    As a result, modern computers are highly parallel and embeds a complex hierarchical caches architecture.
    Writing efficient code for such computers is a complex task.
    Therefore, performance analysis has became an important step of the development of application seeking for performances.

    Most existing performance analysis tools focuses on the point of view of the processor.
    Theses tools see the main memory as a monolithic entity and thus are not able to understand how it is accessed.
    However, memory is a common bottleneck in high performance computing, and the pattern of memory accesses can impact significantly the performances.
    There are a few tools to analyze memory performances, however theses tools are based on a coarse grain sampling.
    Consequently, they focus on a small part of the execution missing the global memory behavior.
    Furthermore, these coarse grain sampling are not able to collect memory accesses patterns.

    In this thesis we propose two different tools to analyze the memory behavior of an application.
    The first tool is designed specifically for Not Uniform Memory Access machines and provides a visualization of the global sharing pattern inside each data structure between the threads.
    The second one collects fine grain memory traces with temporal information.
    We can visualize theses traces either with a generic trace management framework or with a programmatic exploration using R.
    Furthermore we evaluate both of these tools, comparing them with state of the art memory analysis tools in terms of performances, precision and completeness.

\end{chapter}

\begin{chapter}{Résumé}
    \DB{Résumé en Français}
\end{chapter}


\shorttoc{smalltoc}{Outline}{1}

\listoftodos

\mainmatter

\Input{intro}
\Input{profiling}
\Input{mem}
\Input{tabarnac}
\Input{moca}
\Input{analyzing}
\Input{cncl}

\backmatter

\setcounter{tocdepth}{3}
\tableofcontents

\listoffigures
\listoftables
\lstlistoflistings
\printglossary[type=acronym]
\printglossary[type=main]
\bibliographystyle{alpha}
%\bibliographystyle{plain}
\bibliography{biblio}

%\chapter{Acknowledgments}
%\DB{C'est là quj'ai le droit de dire plein de conneries !}
%
\stopcontents[smalltoc]
%==============================================================================

\end{document}
% vim: et si sta lbr  sw=4 ts=4 spelllang=en_us
